Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-17 doc/mc-paper.tex
General updates

\newcommand{\varset}{X}
\newcommand{\exvarset}{Z}
\newcommand{\dependencyset}{{\cal D}}
\newcommand{\litset}{{\cal L}}
\newcommand{\ring}{{\cal R}}
\newcommand{\dset}{{\cal A}}
\newcommand{\rep}{\textbf{R}}
\newcommand{\radd}{+}
\newcommand{\rmul}{\times}
\newcommand{\addident}{\textbf{0}}
\newcommand{\mulident}{\textbf{1}}
\newcommand{\imply}{\Rightarrow}
\newcommand{\ifandonlyif}{\Leftrightarrow}
\newcommand{\drational}{\mathbb{Q}_{2,5}}

@@

\emph{Knowledge compilation} is the process of converting a Boolean
formula into a form that makes it possible to calculate associated
values efficiently, such as the weighted or unweighted model
count. Existing knowledge compilers, however,
provide no guarantee that the representation they generate
is logically equivalent to the original formula, and therefore they cannot validate the generated counts.

We present \emph{Partitioned Operation Graphs} (POGs), a generalization
of representations commonly used in knowledge compilation.
We have designed a proof framework that can prove the equivalence of a
POG representation with a Boolean formula in conjunctive normal form (CNF),
the standard input format for SAT solvers and model counting tools.

We have developed a program that can generate POG representations from
the d-DNNF graphs generated by the state-of-the-art knowledge compiler
\dfour, as well as independently checkable proofs that the output POGs
are equivalent to the input CNF formulas.  We have implemented an
efficient checker that can handle very large proofs. Our tool chain
for generating and verifying POGs scales to all but the largest d-DNNF
graphs generated by \dfour{} for formulas from a recent model counting
competition. Moreover, we present a checker and a model counter for POGs that have
been formally verified in the \lean{} proof assistant, thereby
creating the first formally verified toolchain for
weighted and unweighted
model counting.
\end{abstract}

@@

To further enhance the trustworthiness of our toolchain, we have
developed formally verified versions of the proof checker and ring
evaluator in the \lean{} proof assistant.  Running these two programs
has the effect of formally verifying both the logical equivalence of
the POG to the input formula and the correctness of the count.  In
addition, the existence of a formally verified checker serves the
important role of establishing the soundness of our proof framework.
Indeed, in the course of developing a formally verified checker, we
identified subtle aspects of the proof obligations required to assure
soundness.

@@

  For Boolean formula $\phi$, and ring weight function $w$, the {\em ring evaluation problem} is to compute:
  \begin{eqnarray}
    \rep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
  \end{eqnarray}
  In this equation, sum \scalebox{0.8}{$\sum$} is computed using addition operation $\radd$, and product \scalebox{0.8}{$\prod$} is computed using multiplication operation $\rmul$.
\end{definition}

@@

The {\em weighted model counting} (WMC) problem is also defined over
rational numbers.  Some formulations of WMC allow
independently assigning weights $W(x)$ and $W(\obar{x})$ for each variable $x$ and its complement, with the possibility that
$W(x) + W(\obar{x}) \not = 1$.
We can cast this as a
ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
performing ring evaluation with label $w(x) = W(x)/r(x)$ for each
variable $x$, and computing the weighted count
as $\rep(\phi, w)\; \rmul\; \prod_{x \in \varset} r(x)$.
Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.

The {\em function hashing problem} provides a test
of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, chose a value from $\dset$ at random for $w(x)$.  Two formulas
$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1, w) = \rep(\phi_2, w)$
if they are logically equivalent.  If they are not equivalent, then
the probability that $\rep(\phi_1, w) \not = \rep(\phi_2, w)$ will be at
least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
Function hashing can therefore be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.

@@

\section{Partitioned-Operation Graphs (POGs)}

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
\label{prop:ring:eval}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any labeling $w$:
\begin{eqnarray}
%% V1
\textstyle
\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
\textstyle
\scalebox{1.2}{\rep}\left(\Pand_{i=1,k} \phi_i,\; w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
\textstyle
\scalebox{1.2}{\rep}\left(\Por_{i=1,k} \phi_i,\; w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
%% V2
%\rep(\boolnot \phi, w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
%\rep\left(\phi_1, w \pand \dots \pand \phi_k, w \right) &=& \prod_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:product} \\
%\rep\left(\phi_1, w \por \dots \por \phi_k, w \right) &=& \sum_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:sum}
%\rep\left(\Pand_{i=1,k} \phi_i, w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
%\rep\left(\Por_{i=1,k} \phi_i, w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
\end{eqnarray}
\end{proposition}

@@

\item They allow arbitrary arguments to a $\por$ operation, as long as
  they have disjoint models.  By contrast, each
  sum node $\nodeu$ in a d-DNNF can have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
  any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
%%Our generalization is provided to allow an
%%  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} as
%%  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
\end{itemize}
  Although not required for representing d-DNNF graphs,
  both of these generalizations allow more flexibility in the POG
  representation while maintaining the ability to efficiently perform ring evaluation.

@@

A proof in our framework takes the form of a sequence of clause addition
and deletion steps, with each step maintaining logical equivalence to
the original formula.  More precisely, the status of the proof at any step is represented as
a set of {\em active} clauses $\phi$, i.e., those that
have been added but not yet deleted.  Our framework is based
on {\em extended} resolution~\cite{Tseitin:1983}, meaning that proof
steps can introduce new {\em extension variables} encoding Boolean formulas over input and prior extension variables.
Let $Z$
denote the set of extension variables occuring in formula $\phi$.
Starting with input formula $\inputformula$,
we want to maintain the invariant that
$\inputformula \ifandonlyif \exists Z\,\phi$.

Clauses can be added in two different ways.  One is when they serve as
the {\em defining clauses} for an extension variable.  This form
occurs only when defining $\pand$ and $\por$ operations, as is
described in Section~\ref{sect:crat}.  The other is by {\em
  implication redundancy}.  That is, we can add clause $C$ to the set
of clauses forming formula $\phi$ when $\phi \imply C$.  Clauses can
also be deleted through implication redundancy.  That is, given a set
of clauses $\{C\} \cup \phi$, we can remove $C$ if $\phi \imply C$.

We use {\em reverse unit propagation} (RUP) to certify
implication redundancy when adding or deleting
clauses~\cite{goldberg,vangelder08_verifying_rup_proofs}.
RUP
is the core rule supported by standard
proof checkers~\cite{RAT,wetzler14_drattrim} for propositional logic.
It is a way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
Let $C = \{\lit_1, \lit_2, \ldots,\lit_p\}$ be a clause to be
proved redundant with respect to formula $\phi$.  Let $D_1, D_2, \ldots, D_k$ be a sequence of supporting
{\em antecedent} clauses, such that each $D_i$ is in $\phi$.
A RUP step
proves that $\bigwedge_{1\leq i \leq k} D_i \imply C$ by showing
that the combination of the antecedents plus the negation of $C$ leads
to a contradiction.  The negation of $C$ is the formula
$\overline{\lit}_1 \land \overline{\lit}_2 \land \cdots \land
\overline{\lit}_p$ having a CNF representation consisting of $p$ unit
clauses of the form $\obar{\lit}_i$ for $1 \leq i \leq p$.  A RUP
check processes the clauses of the antecedent in sequence, inferring
additional unit clauses.  In processing clause $D_i$, if all but one
of the literals in the clause is the negation of one of the
accumulated unit clauses, then we can add this literal to the
accumulated set.  That is, all but this literal have been falsified,
and so it must be set to true for the clause to be satisfied.  The
final step with clause $D_k$ must cause a contradiction, i.e., all of
its literals are falsified by the accumulated unit clauses.

Compared to the proofs of unsatisfiability generated by SAT solvers,
it can be seen that ours has important differences.  Most
significantly, we want our proofs to maintain logical equivalence at
each step, whereas an unsatisfiability proof need only guarantee that
no proof step makes a satisfiable set of clauses become
unsatisfiable.  Therefore, our proofs must justify clause deletions as well as clause additions.

@@

The CRAT format draws its inspiration from the LRAT~\cite{lrat} and
QRAT~\cite{heule:JAR2014} formats for unquantified and quantified Boolean formulas, respectively.
Key properties include:
\begin{itemize}
  \item
  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
  Declaring a node $\nodeu$
implicitly adds an {\em extension} variable $u$ and a set of {\em defining} clauses $\theta_{\nodeu}$
  encoding the product or sum operation.
  This is the only means for adding extension variables to the proof.
\item Boolean negation is supported implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\item
  The file contains explicit clause addition steps.
  A clause can only be added if it is logically implied by the existing clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item
  The file contains explicit clause deletion steps.
  A clause can only be deleted if it is logically implied by the remaining clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item The checker must track the dependency set for every input and
  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
  The associated extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.
  Only binary $\por$ operations are allowed.  Generalizing to $k$ arguments would require
  $k\,(k-1)/2$ proofs of disjointness.
\end{itemize}

@@

The sequence of operator declarations, asserted clauses, and
clause deletions represents a systematic transformation of the input formula
into a POG formula.  Validating all of these steps serves to prove that the
POG is logically equivalent to the input formula.
At the completion of the proof, the following conditions must hold:
\begin{enumerate}
\item There is exactly one remaining clause that was added via RUP
  addition, and this is a unit clause consisting of root literal $r$.
\item All of the input clauses have been deleted.
\end{enumerate}
Let $\inputformula$ denote the input formula and $\pogformula$ represent the formula encoded by unit clause $r$ and the defining clauses for the POG nodes:
$\pogformula = \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{\nodeu}$.  We write $\modelset(P)$ as a shorthand for $\modelset(\pogformula)$, recognizing that
any assignment to the input variables implicitly defines the assignments to the extension variables.
The sequence of clause addition steps provide a {\em forward implication} proof that
$\modelset(\inputformula) \subseteq \modelset(P)$.  That is, any total
assignment $\assign$ satisfying the input formula must also satisfy
the formula represented by the POG\@.
Conversely,
each proof step that deletes an input clause proves that any
total assignment $\alpha$ that falsifies the clause must
assign $\assign(r) = 0$.  Deleting all but the final asserted clause and all input clauses provides a {\em reverse implication} proof
that
$\modelset(P) \subseteq \modelset(\inputformula)$.

@@

%%Figure~\ref{fig:chain} illustrates one method of generating a CRAT
%%file from an input formula in conjunctive normal form.  Our tool chain
%%is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
%%but it would also work for other tools that use a top-down search
%%strategy to generate a d-DNNF representation.
Converting a d-DNNF
graph into a POG is straightforward---each node in a d-DNNF graph can
be directly represented as a POG node.  Rather than representing every
d-DNNF node explicitly, our program performs simplifications to
eliminate Boolean constants.
Except in degenerate cases,
%where the formula is unsatisfiable or a tautology,
we can therefore assume
that the POG does not contain any constant nodes.
In addition, negation is only
applied to variables, and so we can merge the negation operations into the terminal nodes,
viewing the POG as consisting
of {\em literal} nodes corresponding to input variables and their negations, along with
{\em terminal} nodes, which can be further classified as {\em product} nodes and {\em sum} nodes.

@@

The monolithic approach
takes advantage of the clausal representations of
the input formula $\inputformula$ and the POG formula $\pogformula$, consisting of the defining clauses of its nodes plus a unit clause for root literal $r$.
Furthermore, it can
negate the POG formula by simply negating its root literal.  More precisely $\neg \pogformula = \bigcup_{\nodeu\in P} \simplify{\theta_{\nodeu}}{\{\obar{r}\}}$
Forward implication
will hold
when $\inputformula \imply \pogformula$.  Equivalently,
the formula $\inputformula \land \neg \pogformula$
should be unsatisfiable, where the
conjunction can be expressed as the union
of the two sets of clauses.  The proof generator writes the clauses to a file and invokes a proof-generating SAT solver.
For each clause $C$ in the unsatisfiability proof, it adds clause $\{r\} \cup C$ to the CRAT proof, and so the empty clause in the proof becomes the unit clause $r$.
Our experimental results show
that this approach can be very effective and generates short proofs
for smaller problems, but it does not scale well enough for routine
use.

We describe the structural approach to proof generation as a recursive procedure
$\validate(\nodeu, \passign, \psi)$ taking as arguments POG
node $\nodeu$, partial assignment
$\passign$, and a set of clauses $\psi \subseteq \inputformula$.
The procedure adds a number of clauses to the proof, culminating with
the addition of the {\em target} clause:
$u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating
that $\left(\bigwedge_{\lit \in \passign} \lit\right) \imply u$, i.e.,
that any total
assignment $\assign$ such that $\passign \subseteq \assign$
will assign $\assign(u) = 1$.
The top-level call has $\nodeu = \noder$, $\passign = \emptyset$, and $\psi = \inputformula$.
The result will therefore be to add unit clause $r$ to the proof.
Here we present a correct, but somewhat inefficient formulation of
$\validate$.  We then refine it with some optimizations.

@@

\section{Experimental Evaluation}
\label{sect:experimental}
We have implemented versions of the three programs that, along with the \dfour{} knowledge compiler,  form the
toolchain illustrated in Figure~\ref{fig:chain}.  One version,
including a proof generator, proof checker, and ring evaluator for
standard and weighted model counting, is designed to be fully
scalable.  The other version, consisting of a proof checker and
ring evaluator has been formally verified within the \lean{} proof
system.
Our long-term goal is to have a toolchain suitable for use in a
knowledge compiler competition, along the lines of the annual
competitions for other automated reasoning tools.  Our experiments
consider the benchmark problems and computing capabilities that would
arise in such a competition and evaluate whether our implementations
could handle these problems.  As with the original proof-generating
SAT competitions, we assume that the proofs at a competition could be
checked with the unverified checker.  Having a verified checker
that works on some subset of the benchmarks would provide an
additional degree of certainty.

@@

Our ring evaluator can perform both standard and weighted model
counting.  It performs arithmetic over a subset of the rationals
consisting of numbers of the form
$a \cdot 2^{b} \cdot 5^{c}$, for integers $a$, $b$, and $c$, and
with $a$ implemented to have arbitrary range.
Allowing scaling by powers of 2 enables the density computation
and rescaling required for standard model counting.  Allowing scaling
by powers of both 2 and 5 enables exact decimal arithmetic, handling
the weights used in the weighted model counting competitions.

@@

Our experiments demonstrate that our toolchain could handle a
substantial portion of the solutions generated in a knowledge compiler
competition.  Further engineering and optimization of our proof
generator and checker could improve their performance and capacity,
and we could increase the time limit for these programs.

>>>>

\newcommand{\varset}{X}
\newcommand{\exvarset}{Z}
\newcommand{\dependencyset}{{\cal D}}
\newcommand{\litset}{{\cal L}}
\newcommand{\ring}{{\cal R}}
\newcommand{\dset}{{\cal A}}
\newcommand{\rep}{\textbf{R}}
\newcommand{\radd}{+}
\newcommand{\rmul}{\times}
\newcommand{\addident}{\textbf{0}}
\newcommand{\mulident}{\textbf{1}}
\newcommand{\imply}{\Rightarrow}
\newcommand{\ifandonlyif}{\Leftrightarrow}
%\newcommand{\drational}{\mathbb{Q}_{2,5}}
\newcommand{\drational}{\textbf{Q}_{2,5}}

@@

\emph{Knowledge compilation} is the process of converting a Boolean
formula into a form that makes it possible to calculate associated
values efficiently, such as the weighted or unweighted model
count. Existing knowledge compilers
provide no guarantee that the representation they generate
correctly represents the original formula, and therefore they cannot validate the generated counts.

We present \emph{Partitioned Operation Graphs} (POGs), a generalization
of the representations used by existing knowledge compilers.
We have designed a proof framework that can prove the equivalence of a
POG representation with a Boolean formula in conjunctive normal form (CNF),
the standard input format for SAT solvers and model counting tools.

We have developed a program that generates POG representations from
the d-DNNF graphs generated by the state-of-the-art knowledge compiler
\dfour, as well as checkable proofs that the output POGs
are equivalent to the input CNF formulas.  We have implemented an
efficient checker that can handle large proofs. Our tool chain
for generating and verifying POGs scales to all but the largest d-DNNF
graphs generated by \dfour{} for formulas from a recent model counting
competition. We also present a checker and a model counter for POGs that have
been formally verified in the \lean{} proof assistant.  These programs both prove the soundness of our proof framework and
create the first formally verified toolchain for weighted and unweighted model counting.
\end{abstract}

@@

To further enhance the trustworthiness of our toolchain, we have
developed formally verified versions of the proof checker and ring
evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.  Running these two programs
has the effect of formally verifying both the logical equivalence of
the POG to the input formula and the correctness of the count.  In
addition, the existence of a formally verified checker serves the
important role of establishing the soundness of our proof framework.
Indeed, in the course of developing a formally verified checker, we
identified subtle aspects of the proof obligations required to assure
soundness.

@@

  For Boolean formula $\phi$, and ring weight function $w$, the {\em ring evaluation problem} is to compute:
  \begin{eqnarray}
    \rep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
  \end{eqnarray}
  In this equation, sum \scalebox{0.8}{$\sum$} is computed using addition operation $\radd$, and product \scalebox{0.8}{$\prod$} is computed using multiplication operation $\rmul$.
\label{def:weight}
\end{definition}

@@

The {\em weighted model counting}  problem is also defined over
rational numbers.  Some formulations  allow
independently assigning weights $W(x)$ and $W(\obar{x})$ for each variable $x$ and its complement, with the possibility that
$W(x) + W(\obar{x}) \not = 1$.
We can cast this as a
ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
performing ring evaluation with label $w(x) = W(x)/r(x)$ for each
variable $x$, and computing the weighted count
as $\rep(\phi, w)\; \rmul\; \prod_{x \in \varset} r(x)$.
Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.

The {\em function hashing problem} provides a test
of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, chose a value from $\dset$ at random for $w(x)$.  Two formulas
$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1, w) = \rep(\phi_2, w)$
if they are logically equivalent.
%, and if $\rep(\phi_1, w) \not = \rep(\phi_2, w)$, then they are clearly inequivalent.
If they are not equivalent, then
the probability that $\rep(\phi_1, w) \not = \rep(\phi_2, w)$ will be at
least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
Function hashing can therefore be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.

@@

\section{Partitioned-Operation Graphs (POGs)}
\label{sect:pog}

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
\label{prop:ring:eval}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any weight function $w$:
\begin{eqnarray}
%% V1
\textstyle
\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
\textstyle
\scalebox{1.2}{\rep}\left(\Pand_{i=1,k} \phi_i,\; w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
\textstyle
\scalebox{1.2}{\rep}\left(\Por_{i=1,k} \phi_i,\; w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
%% V2
%\rep(\boolnot \phi, w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
%\rep\left(\phi_1, w \pand \dots \pand \phi_k, w \right) &=& \prod_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:product} \\
%\rep\left(\phi_1, w \por \dots \por \phi_k, w \right) &=& \sum_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:sum}
%\rep\left(\Pand_{i=1,k} \phi_i, w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
%\rep\left(\Por_{i=1,k} \phi_i, w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
\end{eqnarray}
\end{proposition}
As is described in Section~\ref{subsection:counting}, we have proved these three equations using \lean{}.

@@

\item They allow arbitrary arguments to a $\por$ operation, as long as
  they have disjoint models.  By contrast, each
  sum node $\nodeu$ in a d-DNNF can have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
  any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
%%Our generalization is provided to allow an
%%  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} as
%%  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
\end{itemize}
  These generalizations allow more flexibility in the POG
  representation while maintaining the ability to efficiently perform ring evaluation.

@@

A proof in our framework takes the form of a sequence of clause addition
and deletion steps, with each step preserving the solutions to
the original formula.  The status of the proof at any step is represented as
a set of {\em active} clauses $\phi$, i.e., those that
have been added but not yet deleted.  Our framework is based
on {\em extended} resolution~\cite{Tseitin:1983}, meaning that proof
steps can introduce new {\em extension variables} encoding Boolean formulas over input and prior extension variables.
Let $Z$
denote the set of extension variables occuring in formula $\phi$.
Starting with input formula $\inputformula$,
the proof must maintain the invariant that
$\inputformula \ifandonlyif \exists Z\,\phi$.

Clauses can be added in two different ways.  One is when they serve as
the {\em defining clauses} for an extension variable.  This form
occurs only when defining $\pand$ and $\por$ operations, as is
described in Section~\ref{sect:crat}.  Clauses can also be added or deleted based on
{\em  implication redundancy}.  That is, when clause $C$ satisfies $\phi \imply C$ for some formula $\phi$, then it can either be added to form the set $\phi \cup \{C\}$ or it can be deleted from that set to form $\phi$.

We use {\em reverse unit propagation} (RUP) to certify
implication redundancy when adding or deleting
clauses~\cite{goldberg,vangelder08_verifying_rup_proofs}.
RUP
is the core rule supported by standard
proof checkers~\cite{wetzler14_drattrim,RAT} for propositional logic. It provides a simple and efficient
way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
Let $C = \{\lit_1, \lit_2, \ldots,\lit_p\}$ be a clause to be
proved redundant with respect to formula $\phi$.  Let $D_1, D_2, \ldots, D_k$ be a sequence of supporting
{\em antecedent} clauses, such that each $D_i$ is in $\phi$.
A RUP step
proves that $\bigwedge_{1\leq i \leq k} D_i \imply C$ by showing
that the combination of the antecedents plus the negation of $C$ leads
to a contradiction.  The negation of $C$ is the formula
$\overline{\lit}_1 \land \overline{\lit}_2 \land \cdots \land
\overline{\lit}_p$ having a CNF representation consisting of $p$ unit
clauses of the form $\obar{\lit}_i$ for $1 \leq i \leq p$.  A RUP
check processes the clauses of the antecedent in sequence, inferring
additional unit clauses.  In processing clause $D_i$, if all but one
of the literals in the clause is the negation of one of the
accumulated unit clauses, then we can add this literal to the
accumulated set.  That is, all but this literal have been falsified,
and so it must be set to true for the clause to be satisfied.  The
final step with clause $D_k$ must cause a contradiction, i.e., all of
its literals are falsified by the accumulated unit clauses.

Compared to the proofs of unsatisfiability generated by SAT solvers,
it can be seen that ours has important differences.  Most
significantly, we want our proofs to maintain preserve the set of solutions at
each step, whereas an unsatisfiability proof need only guarantee that
a proof step does not makes a satisfiable set of clauses become
unsatisfiable.  In particular, our proofs must justify clause deletions as well as clause additions.

@@

The CRAT format draws its inspiration from the LRAT~\cite{lrat} and
QRAT~\cite{heule:JAR2014} formats for unquantified and quantified Boolean formulas, respectively.
Key properties include:
\begin{itemize}
  \item
  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
  Declaring a node $\nodeu$
implicitly adds an {\em extension} variable $u$ and a set of {\em defining} clauses $\theta_{u}$
  encoding the product or sum operation.
  This is the only means for adding extension variables to the proof.
\item Boolean negation is supported implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\item
  The file contains explicit clause addition steps.
  A clause can only be added if it is logically implied by the existing clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item
  The file contains explicit clause deletion steps.
  A clause can only be deleted if it is logically implied by the remaining clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item The checker must track the dependency set for every input and
  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
  The associated extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.
  Only binary $\por$ operations are allowed.  Generalizing to $k$ arguments would require
  $k\,(k-1)/2$ proofs of disjointness.
\end{itemize}

@@

The sequence of operator declarations, asserted clauses, and
clause deletions represents a systematic transformation of the input formula
into a POG formula.  Validating all of these steps serves to prove that the
POG is logically equivalent to the input formula.
At the completion of the proof, the following conditions must hold:
\begin{enumerate}
\item There is exactly one remaining clause that was added via RUP
  addition, and this is a unit clause consisting of root literal $r$.
\item All of the input clauses have been deleted.
\end{enumerate}
Let $\inputformula$ denote the input formula and $\pogformula$ represent the formula encoded by unit clause $r$ and the defining clauses for the POG nodes:
$\pogformula = \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$.  We write $\modelset(P)$ as a shorthand for $\modelset(\pogformula)$, recognizing that
any assignment to the input variables implicitly defines the assignments to the extension variables.
The sequence of clause addition steps provide a {\em forward implication} proof that
$\modelset(\inputformula) \subseteq \modelset(P)$.  That is, any total
assignment $\assign$ satisfying the input formula must also satisfy
the formula represented by the POG\@.
Conversely,
each proof step that deletes an input clause proves that any
total assignment $\alpha$ that falsifies the clause must
assign $\assign(r) = 0$.  Deleting all but the final asserted clause and all input clauses provides a {\em reverse implication} proof
that
$\modelset(P) \subseteq \modelset(\inputformula)$.

@@

%%Figure~\ref{fig:chain} illustrates one method of generating a CRAT
%%file from an input formula in conjunctive normal form.  Our tool chain
%%is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
%%but it would also work for other tools that use a top-down search
%%strategy to generate a d-DNNF representation.
Converting a d-DNNF
graph into a POG is straightforward---each node in a d-DNNF graph can
be directly represented as a POG node, although
our program performs simplifications to
eliminate Boolean constants.
Except in degenerate cases,
%where the formula is unsatisfiable or a tautology,
we can therefore assume
that the POG does not contain any constant nodes.
In addition, negation is only
applied to variables, and so we can merge the negation operations into the terminal nodes,
viewing the POG as consisting
of {\em literal} nodes corresponding to input variables and their negations, along with
{\em nonterminal} nodes, which can be further classified as {\em product} nodes and {\em sum} nodes.

@@

The monolithic approach
takes advantage of the clausal representations of
the input formula $\inputformula$ and the POG formula $\pogformula$.
It can
express the negation of $\pogformula$ as $\neg \pogformula = \bigcup_{\nodeu\in P} \simplify{\theta_{u}}{\{\obar{r}\}}$.
Forward implication
will hold
when $\inputformula \imply \pogformula$, or  equivalently,
that the formula $\inputformula \land \neg \pogformula$
is unsatisfiable, where the
conjunction can be expressed as the union
of the two sets of clauses.  The proof generator writes the clauses to a file and invokes a proof-generating SAT solver.
For each clause $C$ in the unsatisfiability proof, it adds clause $\{r\} \cup C$ to the CRAT proof, and so the empty clause in the proof becomes the unit clause $r$.
Our experimental results show
that this approach can be very effective and generates short proofs
for smaller problems, but it does not scale well enough for routine
use.

We describe the structural approach to proof generation as a recursive procedure
$\validate(\nodeu, \passign, \psi)$ taking as arguments POG
node $\nodeu$, partial assignment
$\passign$, and a set of clauses $\psi \subseteq \inputformula$.
The procedure adds a number of clauses to the proof, culminating with
the addition of the {\em target} clause:
$u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating
that $(\bigwedge_{\lit \in \passign} \lit) \imply u$, i.e.,
that any total
assignment $\assign$ such that $\passign \subseteq \assign$
will assign $\assign(u) = 1$.
The top-level call has $\nodeu = \noder$, $\passign = \emptyset$, and $\psi = \inputformula$.
The result will therefore be to add unit clause $r$ to the proof.
Here we present a correct, but somewhat inefficient formulation of
$\validate$.  We then refine it with some optimizations.

@@

\section{Experimental Evaluation}
\label{sect:experimental}
We have implemented versions of the three programs that, along with the \dfour{} knowledge compiler,  form the
toolchain illustrated in Figure~\ref{fig:chain}.  One version,
including a proof generator, proof checker, and ring evaluator for
standard and weighted model counting, is designed to be fully
scalable.  The other version, consisting of a proof checker and
ring evaluator has been formally verified within the \lean{} proof
system.
Our long-term goal is to have a toolchain suitable for use in a
knowledge compiler competition, along the lines of the annual
competitions for other automated reasoning tools.  (The results could be evaluated for correctness
either by having them generate checkable proofs or through randomized equivalence tests via function hashing.)
Our experiments
consider the benchmark problems and computing capabilities that would
arise in such a competition and evaluate whether our implementationed toolchain
could handle these problems.  As with the original proof-generating
SAT competitions, we assume that the proofs at a competition could be
checked with the unverified checker.  Having a verified checker
that works on some subset of the benchmarks would provide an
additional degree of certainty.

@@

Our ring evaluator can perform both standard and weighted model
counting.  It performs arithmetic over a subset of the rationals
we call $\drational$,
consisting of numbers of the form
$a \cdot 2^{b} \cdot 5^{c}$, for integers $a$, $b$, and $c$, and
with $a$ implemented to have arbitrary range.
Allowing scaling by powers of 2 enables the density computation
and rescaling required for standard model counting.  Allowing scaling
by powers of both 2 and 5 enables exact decimal arithmetic, handling
the weights used in the weighted model counting competitions.

@@

We are hopeful that having checkable proofs for knowledge compilers
will allow them to be used in applications where high levels of trust
are required, and that it will provide a useful tool for developers of
knowledge compilers.  Although our current implementation only handles
the outputs of one particular program, we believe it could be adapted
to work with others, as well as with some model counters.

Our experiments demonstrate that our toolchain can already
 handle problems nearly at the limits of current knowledge compilers.
Further engineering and optimization of our proof
generator and checker could improve their performance and capacity substantially.
